---
title: "Assignment7"
output: html_document
date: "2025-10-15"
---


## Given pretest odds
```{r}
pretest_odds <- 1
sens <- 0.8
spec <- 0.10
```

## Likelihood ratios
```{r}
LR_pos <- sens / (1 - spec)
LR_neg <- (1 - sens) / spec
```

## Post-test odds for a NEGATIVE result
```{r}
post_odds_neg <- pretest_odds * LR_neg
``` 

## Convert odds to probability
```{r}
odds_to_prob <- function(odds) odds / (1 + odds)
pretest_prob <- odds_to_prob(pretest_odds)
post_prob_neg <- odds_to_prob(post_odds_neg)
```
## Print results
```{r}
cat(sprintf("Pretest probability: %.4f (%.1f%%)\n", pretest_prob, 100*pretest_prob))
cat(sprintf("LR-: %.4f\n", LR_neg))
cat(sprintf("Post-test odds (negative): %.4f\n", post_odds_neg))
cat(sprintf("Post-test probability (negative): %.4f (%.1f%%)\n\n", post_prob_neg, 100*post_prob_neg))
```
#Output: Pretest probability: 0.5000 (50.0%)
#LR-: 2.0000
#Post-test odds (negative): 2.0000
#Post-test probability (negative): 0.6667 (66.7%)

## Quality check with a hypothetical pool of 10,000 people
```{r}
N <- 10000
prev <- pretest_prob               # 0.5
D   <- N * prev                    # diseased
H   <- N - D                       # healthy
```


# Test results
```{r}
TN <- H * spec                     # true negatives among healthy
FN <- D * (1 - sens)               # false negatives among diseased
Neg_total <- TN + FN
```

# Probability(disease/negative) from counts
```{r}
post_prob_neg_empirical <- FN / Neg_total

cat(sprintf("Empirical P(disease | negative) from 10,000 cohort: %.4f (%.1f%%)\n",
            post_prob_neg_empirical, 100*post_prob_neg_empirical))
```

#Output: P is 0.6667 (66.7%)

#I will attempt to answer the question using Bayesian uncertainty principle to see if it is the same answer

#Setup

```{r}
set.seed(123)
```

```{r}
prev_mean <- 0.5
Se_mean   <- 0.8
Sp_mean   <- 0.10
```

# Moderate-strength priors 
```{r}
prev_a <- 50; prev_b <- 50       # centers at 0.5
Se_a   <- 16;  Se_b   <- 4       # centers at 0.8
Sp_a   <- 1;   Sp_b   <- 9  
```

# Monte Carlo to propagate parameter uncertainty
```{r}
S <- 100000
prev  <- rbeta(S, prev_a, prev_b)
Se    <- rbeta(S, Se_a,   Se_b)
Sp    <- rbeta(S, Sp_a,   Sp_b)
```

```{r}
p_neg_given_D    <- 1 - Se
p_neg_given_notD <- Sp

p_D_given_neg_draws <- (p_neg_given_D * prev) /
  (p_neg_given_D * prev + p_neg_given_notD * (1 - prev))
```

# Summary
```{r}
mean_p  <- mean(p_D_given_neg_draws)
ci95    <- quantile(p_D_given_neg_draws, c(0.025, 0.975))
list(
  mean_P_D_given_negative = mean_p,
  CI_95 = ci95
)
```
#Output: $mean_P_D_given_negative
#[1] 0.6925907
#$CI_95
# 2.5%     97.5% 
#0.2659651 0.9862343 

#Written Answer:

#The pretest probability of disease was 0.5, which corresponds to pretest odds of 1 (a 50% chance of disease before testing). Using the given sensitivity of 0.8 and specificity of 0.10, the likelihood ratio for a negative test was calculated as (1−0.8)/0.10=2. The post-test odds are found by multiplying the pretest odds by this likelihood ratio, giving 1×2=2. Converting odds back to probability yields 2/(1+2)=0.67, or approximately a 66.7% chance of disease after a negative test. This result is paradoxical because, despite a negative result, the probability of disease actually increases from 50% to 66.7%. The low specificity of the test (only 10%) means that true negatives are rare, so a negative result is more likely to come from someone who actually has the disease. Therefore, the post-test odds of disease after a negative test are 2, showing that this diagnostic test is misleading and would not be clinically useful without considering its likelihood ratios beforehand.

#To further examine how uncertainty in test performance could affect results, I performed a Bayesian uncertainty analysis priors centered on the given parameters (prevalence = 0.5, sensitivity = 0.8, specificity = 0.10). The simulation drew 100,000 samples from these distributions to propagate uncertainty through the Bayes formula. The resulting posterior mean for P(Disease/Negative) was approximately 0.693, with a 95% credible interval ranging from 0.266 to 0.986. This interval indicates that even when accounting for variability in the input parameters, the expected probability of disease after a negative test remains close to the deterministic estimate of 0.667. However, the wide credible interval reflects how sensitive the post-test probability is to uncertainty in specificity, which was modeled with a relatively weak prior. Overall, the Bayesian approach reinforces the paradoxical finding that a negative test result still implies a high likelihood of disease, highlighting that the test’s low specificity severely limits its diagnostic capabilities.



