---
title: ''
author: "Jonathan Flowers"
date: "2025-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Population Genetics and Evolutionary Biology for Bioinformatics Assignment 1

The field evolutionary biology depends on phylogeny to understand the evolutionary history of life on earth. Increasingly, phylogeny is used in other fields including epidemiology, comparative genomics, and medical genomics including cancer biology, where single cell phylogenetics has become an important tool to understand tumor progression (Schwarz and Schaeffer 2017 Nature Reviews Genetics doi: 10.1038/nrg.2016.170). 

This assignment introduces practical methods to phylogeny construction using the Phangorn package in R. The assignment provides a hands opportunity to create a multiple sequence alignment, perform both distance-based and Maximum Likelihood approaches to tree-building, and introduce the most common approach to assess confidence in phylogenitic relationships. 

### About the data

The data are from Uddin et al. (2008) phylogenetic study of great apes, new and old world monkeys (doi: 10.1186/1471-2148-8-8) using the COX5A gene. The COX5A sequences are in FASTA format and have not been aligned.

### Completing your assignment

This assignment is provided as an .Rmd file which can be open in RStudio and code chunks run interactively. 

Please return an updated .Rmd and a knitted .html file, both with your answers. Please include your answers in **bold** and figures included after the appropriate questions. You can use this .Rmd document as a template. 

If for some reason you are unable to return a knitted report, please include your report in a single pdf or word document.

On the NYU Brightspace website, you will find an entry for this assignment in the Assignment section. Please upload your completed assignment at the provided link. 

**Please include your name in the filenames and inside the file of any uploads to Brightspace.**

### General instructions for completing your assignment

For all tasks in this assignment you may wish to review materials from Weeks 1 and 2 and the code review from Week 2 (See code_review1.phylogeny_reconstruction.Rmd).

Before beginning, you need to install the phangorn and tidyverse packages (available at CRAN repository) which can be installed with the familiar `install.packages("package name here")` syntax from your RStudio console. The required `msa`, `ggtree` and `treeio` packages can be installed by following instructions at Bioconductor.

msa:
<https://bioconductor.org/packages/release/bioc/html/msa.html>

ggtree:
<https://bioconductor.org/packages/release/bioc/html/ggtree.html>

treeio: 
<https://bioconductor.org/packages/release/bioc/html/treeio.html>

Data:
Download the raw unaligned COX5A sequences in fasta format from the same Google Drive directory as you obtained the .Rmd (location provided in Brightspace)

**The preferred way to complete the assignment is to first run the code interactively via RStudio (by clicking the green arrow in each code block beginning at the top of the .Rmd), fill in your answers (you can help the TA by putting your answers in bold), and then knit your document into an .html file (See Live Session from 9/13/23 and example.Rmd presented in that session) to create a final report.**

### Task 1: Introduction to phylogeny interpretation

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("msa")
```
```{r}
BiocManager::install("ggtree")
```
```{r}
BiocManager::install("treeio")
```

Execute the following code and answers the following questions about the trees.

```{r}
library(tidyverse)
library(ape)
library(phangorn)
library(ggtree)
library(treeio)
library(msa)
```
```{r}
treeA.phylo <- ape::read.tree(text = '(((A:1,B:1):1,C:2):1,D:3);')
treeB.phylo <- ape::read.tree(text = '((A:1,B:1):2,(C:1,D:1):2);')
treeC.phylo <- ape::read.tree(text = '(((B:1,A:1):1,D:2):1,C:3);')
treeD.phylo <- ape::read.tree(text = '(C:3,((A:1,B:1):1,D:2):1);')
tree.multiPhylo <- list(treeA.phylo,treeB.phylo,treeC.phylo,treeD.phylo)
class(tree.multiPhylo) <- 'multiPhylo'

tree.multiPhylo %>%
ggtree() +
  geom_tree() +
  geom_tiplab() +
  facet_wrap(~ .id) 
```

**Q1.1** If the four trees are unrooted and branch lengths not drawn to scale, which statement is correct? Select the single best answer (1 point).

(a) All for trees are equivalent
(b) All for trees are not equivalent
(c) Only tree 3 and 4 are equivalent
(d) Only tree 1 and 3 are equivalent
(e) Both c and d

**a**

**Q1.2** If Tree 1 is the optimum tree in an analysis where taxon A was chosen a priori to be the outgroup taxon, then which of the following statements is correct? Select the single best answer (1 point)

(a) B and C form a monophyletic group
(b) B and D form a monophyletic group 
(c) C and D form a monophyletic group
(d) A and B form a monophyletic group

**c**
 
**Q1.3** Now assume the four trees are rooted with the root edge at the left-most vertical edge of the tree. Which of the following statements is correct. Select the single best answer (1 point).

(a) All for trees are equivalent
(b) All for trees are not equivalent
(c) Only tree 3 and 4 are equivalent
(d) Only tree 1 and 3 are equivalent
(e) Both c and d

**c**

```{r}

tree.phylo <- ape::read.tree(text = '(((A:1,B:1):2,(C:1,D:1):2),E:3);' )


tree.phylo %>%
ggtree() +
  geom_tree() +
  geom_tiplab()
```
**Q1.4** Assume the above tree is rooted using an outgroup taxon E. If the node joining A and B as sister taxa is found to have low boostrap support and the node joining C and D has high bootstrap support, then which of the following statements are true? Select all that apply (1 point)

(a) taxon C is more closely related to A or B than to taxon D
(b) taxon D is more closely related to A or B than to taxon C
(c) taxon A and B form a monophyletic group
(d) taxon C and D form a monophyletic group
(e) the taxa A, B, and the clade consisting of C and D form a polytomy

 **d and e**
 
### Task 2: Create a multiple sequence alignment

Use the example code provided in the Week 2 code review (code_review1.phylogeny_reconstruction.Rmd) to read the un-aligned COX5A fasta file into your .Rmd file then create an msa using the MUSCLE aligner. The output of your code should produce an object of class MsaDNAMultipleAlignment for input into the next step. Subsequent steps will work seemlessly if your alignment objects is assigned to variable named `msa.MsaDNAMultipleAlignment`.

**My code**

# Install package
```{r}
BiocManager::install("Biostrings")
```

# Path to file
```{r}
cox5a_fasta<- "~/Documents/Population Genetics and Evolutionary Biology for Bioinformaticians/COX5A_unaligned.fa"
```


# Read unaligned sequences as a DNAStringSet
```{r}
seq.DNAStringSet <- readDNAStringSet(cox5a_fasta[1])
seq.DNAStringSet  #check
```


# Create a multiple sequence alignment with MUSCLE (dna mode)
```{r}
msa.MsaDNAMultipleAlignment <- msa(seq.DNAStringSet,
                                    type   = "dna",
                                    method = "Muscle"   # MUSCLE aligner via msa
)
```


# Confirm class
```{r}
class(msa.MsaDNAMultipleAlignment)
```
**Output "MsaDNAMultipleAlignment" attr(,"package") [1] "msa"**

```{r}
# read unaligned D-loop sequences of dog breeds and their relatives
seq.DNAStringSet <- readDNAStringSet('COX5A_unaligned.fa')
seq.DNAStringSet

# confirm the type of object
class(seq.DNAStringSet)
```

```{r}
# Create a multiple sequence alignment using the Muscle aligner
msa.MsaDNAMultipleAlignment <- msa(seq.DNAStringSet,
                                   type='dna',
                                   method='Muscle')
msa.MsaDNAMultipleAlignment
class(msa.MsaDNAMultipleAlignment)
```

```{r}
# Convert between objects of class MsaDNAMultipleAlignment (msa package) to phyDat (Phangorn) and DNAbin (ape) object classes. DNAbin is a widely interoperable object class among packages in R, while the others are not.
# note: the msa package has a helpful set of conversion functions to more widely used object classes for multiple sequence alignments
msa.phyDat <- msaConvert(x = msa.MsaDNAMultipleAlignment,
                         type=c("phangorn::phyDat"))
msa.DNAbin <- msaConvert(x = msa.MsaDNAMultipleAlignment,
                         type = c("ape::DNAbin"))

class(msa.phyDat)
class(msa.DNAbin)
```

Visualize the MSA. The alignment should not have an excessive number of gaps (from insertion/deletion mutations) and poorly aligned regions. "Gappy" alignments need to be "cleaned" to produce an alignment containing only columns in the msa that are clearly homologous. Cleaning alignments to remove ambiguously aligned sites is a critical step to any phylogenetic study, particularly when comparing divergent taxa. In fact, obtaining high quality MSAs is one of the challenges of automating the alignment process at a genome scale. Tools like GBlocks <https://doi.org/10.1093/oxfordjournals.molbev.a026334> are helpful in this context.

#formatted alignment

```{r}
msa.MsaDNAMultipleAlignment <- msa(seq.DNAStringSet,
                                   type   = "dna",
                                   method = "Muscle")
```
```{r}
msaPrettyPrint(msa.MsaDNAMultipleAlignment, output="asis")
```

```{r}
# note: the variable, msa.MsaDNAMultipleAlignment corresponds to the alignment object output by the msa() function. Adjust the variable name accoridingly if you used a different name
print(msa.MsaDNAMultipleAlignment, show="complete")
```

**Q2.1.** Review the alignment and answer the following.

How many insertion/deletions are in your alignment (2 points)?
(a) 1  
(b) 2  
(c) 6  
(d) 10  

**B**

### Task 3: Build a distance-based phylogeny

You are now ready to build a phylogenetic tree from the MSA using a distance method. Trees constructed in this fashion are similar to hierarchical clustering and use the same algorithms as non-evolutionary applications. However, distance-based phylogeny reconstruction typically uses an evolutionary rate model to adjust for known biases as we shall see.

Building a distance tree requires a distance matrix with evolutionary distance estimates between all pairs of taxa in the analysis. In distance-based tree reconstruction, it is often necessary to apply some form of correction to the distance estimate to account for the possiblity of 'multiple hits' (see lecture material), rate differences between transition and tranversion mutations and other factors. Here, we will use the Jukes-Cantor (JC69) correction.

Create a distance matrix using the JC69 model.

```{r}
jc.dist <- dist.dna(msa.DNAbin,model="JC69")
jc.dist
```

Now lets see how much the JC69 distances compare to p-distances. P-distances are the standard hamming distance, or, the proportion of sites that differ between a pair of sequences.

```{r}
p.dist <- dist.dna(msa.DNAbin,model = "raw")
p.dist
```

**Q3.1.** Review the two distance matrices. There are two pairs of sequences that are identical. Which are they? (2 points)  
 **Homo_sapiens76 and Homo_sapiens94 and Pan_paniscus and Pan_troglodytes**

Let's plot the raw p-distances (Hamming distances) compared to the Jukes-Cantor distances. This takes a little data wrangling of the pairwise distance matrices, before we can generate the plot

```{r}

# create a tibble species pairs and 
jc.tbl_df <- as.matrix(jc.dist) %>%
  as.data.frame() %>%
  rownames_to_column(var = 'row_spp') %>%
  pivot_longer(cols = -row_spp,
               names_to = 'col_spp',
               values_to = 'jc_dist') %>%
  filter(row_spp != col_spp) %>%
  mutate(spp_pair = paste(row_spp,col_spp,sep="_"))
jc.tbl_df # view

p.tbl_df <- as.matrix(p.dist) %>%
  as.data.frame() %>%
  rownames_to_column(var = 'row_spp') %>%
  pivot_longer(cols = -row_spp,
               names_to = 'col_spp',
               values_to = 'p_dist') %>%
  filter(row_spp != col_spp) %>%
  mutate(spp_pair = paste(row_spp,col_spp,sep="_"))
p.tbl_df # view
```

Now, we can merge the two tibble with JC and p-distances on species pair 
```{r}
dist.tbl_df <- inner_join(jc.tbl_df,p.tbl_df,by='spp_pair')
dist.tbl_df 
```

Plot the two distance metrics with a line with slope 1 for visualization purposes.
```{r}
dist.tbl_df %>%
  ggplot(aes(x = jc_dist,y = p_dist )) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1)
```

**Q3.2** Review the plot above. Which of the following best explains the difference between the two distance measures? Select all that apply (3 points)

(a) the JC69 distance corrects for multiple hits, p-distances do not  
(b) the JC69 distance includes indels, p-distances do not 
(c) the JC69 distance adjusts for alignment errors, the p-distance does not
(d) the JC69 distance adjusts for sequencing errors, the p-distance does not
(e) JC69 distances under-estimates the evolutionary distance between diverged sequences
(f) p-distances under-estimates the evolutionary distance between diverged sequences

**A and F**


Now let's construct a phylogenetic tree using the Neighbor-joining (NJ) tree with the JC69 distance matrix

```{r}
msa.NJ.phylo <- NJ(jc.dist)

# Sometimes the NJ algorithm produces negative branch lengths, which can be safely ignored to avoid plotting errors
options('ignore.negative.edge' = TRUE)

# Now plot the tree
msa.NJ.phylo %>%
  ggtree() +
  geom_tree(layout = 'dendrogram') +
  geom_tiplab(size = 2) +
  ggtitle("NJ (unrooted)")
```

Phylogenetic trees are not typically rooted by default. Let's confirme the NJ tree above is unrooted.

```{r}
ape::is.rooted(msa.NJ.phylo)
```

Let's try rooting the tree on the outgroup. Lemurs are known outgroups to all species of old world monkeys, gibbons, and great apes, which constitute the rest of the taxa in this analysis. Therefore, we can specify the Lemur (Eulemur fulvus) as the outgroup. The root  will be placed at node that joins the outgroup to ingroup species (?ape::root).

```{r}
msa.NJ.phylo <- root(phy = msa.NJ.phylo, outgroup = "Eulemur_fulvus", resolve.root = T)

msa.NJ.phylo %>%
  ggtree() +
  geom_tree(layout = 'dendrogram') +
  geom_tiplab(size = 2) +
  ggtitle("NJ (rooted)")
```

It is important to take care when interpreting phylogenetic trees, and to first identify if the tree is rooted or not. One way to do this is to identify if an outgroup was included in the study. The ingroup species in a study are those that share a most recent common ancestor (MRCA) in the more recent past than they do with the outgroup species.

In the phylogeny above, Eulemur is the outgroup. Therefore, it is clear that the MRCA of all species in the study is somewhere along the long branch to Eulemur and the MRCA of the ingroup taxa is therefore at the base of Papio and Colobus Old World Monkeys and the other non-Lemur taxa. 

So why isn't the root shown along the branch to Eulemur. The short answer is because the position along this edge is unknown. This issue is discussed by the developer of the ape package here: <https://github.com/emmanuelparadis/ape/issues/143>

When trying to interpret a tree, it is important to first determine if a tree is unrooted or rooted. If it is rooted, if the outgroup is shown then its clear where the root of the tree is. If it is not, then the tree should have an explicit root edge or otherwise indicate the location of the root. In some cases, the outgroup taxon is dropped altogether from the phylogeny after the MRCA of the ingroup taxa has been been identified using the outgroup to root the tree. Unfortunately, these essential details are often not made explicit even in the published literature. In our case, however, there isn't a polytomy at the base of the tree!

Now create a bootstrap tree with bootstrap support values on each branch using the `bootstrap.phyDat` function. Note that we specify argument `bs=100` to generate 100 pseudoreplicate MSAs and trees by bootstrapping over columns in the MSA and that we specify the same NJ clustering method and JC69 model of sequence evolution as we did when generating the original NJ tree.

```{r}
msa.NJ.multiphylo <- bootstrap.phyDat(msa.phyDat, bs = 1000, function(x) NJ(dist.ml(msa.phyDat,model="JC69")))
plotBS(msa.NJ.phylo,msa.NJ.multiphylo,type = "phylogram",main="Neighbor-Joining Bootstrap Tree (unrooted)",cex = 0.5)
```

**Q3.3.** Which of the following best describes the purpose of bootstrapping in phylogeny reconstruction? Selecet the single best answer (2 points)
(a) Bootstrapping is used to estimate confidence intervals around estimates of the rate of evolution on each branch of a tree
(b) Bootstrapping is used to determine the outgroup taxon
(c) Bootstrapping is used to improve efficiency in identify the optimal tree in the universe of possible trees
(d) Bootstrapping is used to assess the level of support for a branch in a phylogeny

**d**

**3.4.** Considering both the location of the root and the bootstrap values, which of the following do not represent a monophyletic group, or clade? (2 points)
(a) Papio anubis and Colobus guereza
(b) Pongo pygmaeus and Nomascus gabriellae
(c) Nomascus gabriellae and Symphalangus_syndactylus
(d) Homo_sapiens94 and Homo_sapiens76

**B**

**Q3.5.** Bootstrapping in molecular phylogeny reconstruction involves sampling columns in the MSA with replacement. Which of the following statements is true about this bootstrap procedure and output tree. Select all true statements. (3 points)  

(a) Each 'pseudoreplicate' MSA has the same number of characters (i.e., columns) as the original MSA  
(b) No column from the original MSA can appear more than once in a pseudoreplicate MSA  
(c) A distance matrix is created for each pseudoreplicate MSA followed by tree construction for that   pseudoreplicate
(d) The bootstrap value on each branch of the bootstrap tree represents the percent of pseudoreplicate trees in which that branch was observed.  
**A,C,D**

Some distance-based algorithms are 'ultrametric' and assume equal rates of evolution (a 'molecular clock'). Such trees can be recognized because branch lengths of taxa that share a most recent common ancestor (MRCA) will have the same branch lengths leading back to the shared ancestor. In non-ultrametric trees, taxa that share an MRCA **may** have different branch lengths, or not. Since such taxa share a common ancestor at the same time in the past, any differences in branch lengths leading to the common ancestor reflects differences in rates of evolution (differences in the rate of accumulation of mutations) in the two species. 

**Q3.6.** Considering the location of the root on the bootstrap tree, list a pair of terminal taxa that form a clade (a monophyletic group), clearly have different rates of evolution, and indicate which species is evolving faster at COX5A (2 points).

**Papio anubis and Colobus guereza, because Papio anubis has a longer terminal branch than Colobus from their MRCA.**


Now use the example above to create a bootstrap tree with Kimura 2-parameter model ("K80") of sequence evolution.

**Q3.7** Show your code and phylogeny (2 points)

# Using phangorn's ML distance under K80:

```{r}
library(ape)
library(phangorn)
library(ggtree)
```
```{r}
msa.phyDat <- msaConvert(msa.MsaDNAMultipleAlignment, type = "phangorn::phyDat")
msa.DNAbin <- msaConvert(msa.MsaDNAMultipleAlignment, type = "ape::DNAbin")
```

```{r}
k80.dist <- dist.dna(msa.DNAbin, model = "K80")
k80.nj   <- NJ(k80.dist)
```

# Root on outgroup
```{r}
k80.nj.root <- root(k80.nj, outgroup = "Eulemur_fulvus", resolve.root = TRUE)
```

```{r}
# --- Bootstrap under K80 ---
set.seed(123)
k80.bs <- bootstrap.phyDat(
  msa.phyDat,
  bs  = 1000,
  FUN = function(x) {
    x_dnabin <- as.DNAbin(x)             # convert the resampled phyDat -> DNAbin
    NJ(dist.dna(x_dnabin, model = "K80"))
  }
)
```

```{r}
plotBS(k80.nj.root, k80.bs, type = "phylogram",
       main = "Neighbor-Joining Bootstrap Tree (K80)", cex = 0.7)
```


Answer:
```{r}
K80.dist <- dist.dna(msa.DNAbin,model = "K80")
msa.NJ.K80.phylo <- NJ(K80.dist)

# Sometimes the NJ algorithm produces negative branch lengths, which can be safely ignored to avoid plotting errors
options('ignore.negative.edge' = TRUE)

# Now plot the tree
msa.NJ.K80.phylo %>%
  ggtree() +
  geom_tree(layout = 'dendrogram') +
  geom_tiplab(size = 2) +
  ggtitle("NJ (unrooted)")
```
```{r}
msa.NJ.K80.phylo <- root(phy = msa.NJ.K80.phylo, outgroup = "Eulemur_fulvus", resolve.root = T)

msa.NJ.K80.phylo %>%
  ggtree() +
  geom_tree(layout = 'dendrogram') +
  geom_tiplab(size = 2) +
  ggtitle("NJ (rooted)")
```

**Q3.8.** Do you notice anything different in the phylogeny? If yes, describe the difference from J69 model (1 point).

**There is no major change. The difference is in branch lengths (slightly re-scaled) because K80 accounts for different rates of transitions vs. transversions, whereas JC69 does not**

**Q3.9.** What is the same and what differs between the J69 and K80 models? Select all true statements (2 points)
(a) JC69 accounts for multiple hits, K80 doesn't
(b) K80 accounts for multiple hits, J69 doesn't
(c) both J69 and K80 account for multiple hits
(d) JC69 has different rate categories for transition and transversion mutations, K80 doesn't
(e) K80 has different rate categories for transition and transversion mutations, J69 doesn't
(f) both J69 and K80 have different rate categories for transition and transversion mutations
(g) both allow for differences in base content

**C and E**


### Task 4: Maximum Likelihood (ML) analysis of COX5A.

Once you have an appropriately curated MSA, the first step in ML-based phylogeny reconstruction is typically to identify a model of sequence evolution that best fits the data. **Some methods like the popular RaxML software assume a General Time Reversible model is best, which may or may not be true**, but for large genome-scale datasets this is sometimes assumed for computational ease. For smaller datasets, the best fit model should be determined with Model Test (or jModelTest), which calculates the Akaike Information Criteria (AIC) and the Bayesian Information Criterion for a large number of substitution models. These criteria can then be used to select the best model of sequence evolution to explain the patterns of mutations in the MSA. 

Note, 'substitution' and 'mutation' are sometimes used loosely as equivalent in the phylogenetics vernacular, although we will see later in the course that substitution and mutation are different processes in molecular evolution. 

Run ModelTest here as implemented in the phangorn package.

```{r}
set.seed(234135)

# Model selection
# note: mt is a "modelTest" data.frame with LogLik, AIC, BIC, etc for each model
mt <- modelTest(msa.phyDat)
```

Now review the output

```{r}
mt %>%
  as_tibble
```

How do we identify the model with the best fit? The Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC) are approaches to identify the best model. Review the AIC and BIC wikipedia pages here <https://en.wikipedia.org/wiki/Akaike_information_criterion> and here <https://en.wikipedia.org/wiki/Bayesian_information_criterion>. 

**Q4.1.** Which model is the best fit according to AIC?  Hint: sort the tibble using the `arrange` verb, either in ascending or descending order to identify the best fit model. (2 points).

(a) GTR+G(4)+I
(b) F81
(c) K80+I
(d) TPM3+I

**A**


Now we can run the ML analysis to determine the tree topology and branch lengths that maximize the probability of observing the sequences in the MSA. Note that we simply provide the Model Test object (mt) to the ML function pml_bb() and it will identify the best model without user input.

```{r}
# note: here we provide a modelTest object instead of alignment
pml_bb.modelTest <- pml_bb(mt, control = pml.control(trace = 0), rearrangement = "stochastic")  # pml_bb.modelTest is a pml object
```


```{r}
?pml_bb
```
**Q4.2.** Review the pml_bb documentation "Details" (at your console, enter `?pml_bb`). What model selection method does this function use to select the best-fitting model of sequence evolution from the modelTest result object? (2 points)
(a) AIC
(b) BIC
(c) Maximum Likelihood
(c
**B**

Once again, check if the tree is rooted
```{r}
# note: the tree is found "tree" slot of pml object and is an object of class phylo
pml_bb.modelTest$tree %>%
  ape::is.rooted()
```

Lets root using same approach as with NJ distance tree and plot the ML tree.

```{r}
# note: we can plot a pml object with base R plot function or convert to treedata object and plot with ggtree
pml_bb.modelTest$tree %>% 
  ape::root(outgroup = "Eulemur_fulvus", edgelabel = TRUE,resolve.root=T) %>%
  as_tibble() %>%
  as.treedata() %>%
  ggtree(layout="rectangular") + 
    geom_tiplab()
```

**Q4.3** Take a screenshot of your ML tree and upload along with your completed assignment. **This is required because in principle, when you knit your Rmd into an .html report, your second run of the ML search algorithm may identify a different ML tree (2 points)

**Q4.4.** Now compare your ML tree, to your JC69 NJ tree in Task 2. If we take at face value the branching pattern in each analysis, which phylogeny correctly shows monkeys and all great apes and gibbons sharing a most recent common ancestor? (2 points)?

**Eulemur is the outgroup and the remaining taxa— (Papio, Colobus), gibbons (Nomascus, Symphalangus), and great apes (human, chimp, bonobo, gorilla, orangutan)—form a single clade (Catarrhini). Both phylogenies show monkeys and all great apes**

Optimality criterion methods like ML search the "tree universe" (either evaluating all of the tree topologies in an exhaustive search or a subset of possible tree topologies using a tree-searching algorithm) to identify the tree that maximizes the optimality score, in this case, the Maximum Likelihood.  

Ideally, we would conduct an exhaustive search of all trees in the tree universe to determine their likelihood, thereby ensuring that we infer the ML tree and do not end up with a suboptimal tree (which can occur if the tree-searching algorithm gets stuck in a local optimum). Instead, we applied a stochastic tree-searching algorithm using a stochastic search strategy. This approach uses the stochastic algorithm to search a subset of trees, but this is not guarenteed to obtain the ML tree. 

**Q4.5.** Review the Google Slides from Week 2. What is the total number of rooted trees that exist in the tree universe for the MSA in this assignment? (2 points)
**(2n−3)!!=(2⋅11−3)!!=19!!=19×17×15×13×11×9×7×5×3×1=654,729,075**



